# Distribution map using survey data
## An example of Starry ray in the North Sea, using INLA
#### Authors: Geert Aarts, Niels Hintzen, Harriet van Overzee, Jurgen Batsleer, Jan Jaap Poos, Ingrid Tulp

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 110)
rm(list=ls())
```

Before we get started we need to set the correct path for accessing the data, and to load the relevant libraries. The INLA package can be installed using install.packages("INLA", repos="https://inla.r-inla-download.org/R/stable"). We need the inla package, but also a number of packages for spatial analyses ('rgeos' and 'rgdal')  and for plotting the data and the results ('lattice', 'latticeExtra', 'grid','gridExtra','mapdata', and 'maptools').

```{r, eval=T, results="hide", echo=TRUE, message=FALSE, warning=FALSE}
path <- "~/WMR-INLA/data"
#path <- "d://WMR-INLA/data" 

library(INLA); library(fields); 
library(mgcv)
library(lattice); library(latticeExtra); library(grid); library(gridExtra);
library(rgdal); library(rgeos); 
library(mapdata); library(maptools)

load(file.path(path,"INLA_results.Rdata"))
```

## Simulate data by sampling from posterior (from I4nb)

Next we want to simulate a number realizations and integrate over surface (per year) so that we get a population level estimate.  

```{r, eval=T, echo=T}
# Simulate regression parameters using inla.posterior.sample 
# to simulate from the model. The output is stored in the Sim object.
set.seed(1234)
NSim <- 1000
Sim <- inla.posterior.sample(n = NSim, result = I4nb)
```

In this example, we thus sample `r NSim` times. The Sim object is a list of length NSim. Each element of  this list contains a single realization. Let's have a closer look.

```{r, eval=T, echo=T}
names(Sim[[1]])

#get out names of different types of rows in latent
rnames <- rownames(Sim[[1]]$latent)
rtypes <- unique(unlist(lapply(strsplit(rnames,":"),function(x){x[1]})))
rtypes
```
Now, there are a number of different types of rows in latent: APredictor, Predictor, W, Depth, and the fixed effects.

```{r, eval=T, echo=T}
#to get the w par realizations out: 
rtypes[3]
wrownum <- grep(paste0("^",rtypes[3]),rownames(Sim[[1]]$latent))
wmat <- sapply(Sim, function(x) {x$latent[wrownum]})
dim(wmat)

#to get the depth realizations out: 
rtypes[4]
drownum <- grep(paste0("^",rtypes[4]),rownames(Sim[[1]]$latent))
dmat <- sapply(Sim, function(x) {x$latent[drownum]})
dim(dmat)

#to get the fixed effect par realizations out (1:4 now because depth was added) 
fixed <- rtypes[-(1:4)]
lrownum <- unlist(lapply(fixed, function(x) {grep(x, rownames(Sim[[1]]$latent), fixed = TRUE)}    ))
linmat <- sapply(Sim, function(x) {x$latent[lrownum]})
dim(linmat)
dimnames(linmat)[[1]] <- fixed
```
The second dimensions of wmat, dmat and linmat are equal to the number of simulations, here `r NSim`. The first dimension of wmat is equal to ... . The first dimension of dmat is equal to the number of depths in the original dataset, see for instance length(unique(cpue_subset$Depth)). We do not have all depths in the depth range in our original dataset, so some interpolation is required.

We'll use the approx function to do a linear interpolation using the approx function. In that function we can specify the x and y values for which we have information, and estimate the y values at each depth in the grid using the xout argument.

First, we'll need the depth at each grid cell. The locations are in the wproj object as wproj$x and wproj$y. However, the x and y in the gam model that we use are called Xkm and Ykm. Thus, we need some translation first.

A quick plot of the depths (with predictions between 0 and 250 m reveals that the depths are OK. Check if this is still the case when plotting the depths object (that is a matrix with sizes equal to wproj)
```{r, eval=T, echo=T}

fullgrid <- expand.grid(Xkm=unique(wproj$x), Ykm = unique(wproj$y))

fullgrid$depth <- predict(depthmodel, newdata=fullgrid) 
levelplot(depth ~ Xkm * Ykm,  data = fullgrid, at=seq(0,250,10))

depths <- matrix(fullgrid$depth, nrow=length(wproj$x),ncol=length(wproj$y), dimnames = list("x" = wproj$x, "y" = wproj$y))

#constrain depths to be within min and max observed depths (otherwise interpolation will fail later)
depths <- apply(depths, c(1, 2), function(x) min(max(x,min(cpue_subset$Depth)),max(cpue_subset$Depth)))

#take only depths where our grid is (by using info from earlier plot)
depths[is.na(w.pm1km2)] <- NA 
image(depths)
```

length(wm.pm1km2[!is.na(wm.pm1km2)]) contains surface of grid in km2, because it was generated as a 1 by 1 km grid. 

```{r, eval=T, echo=T}

numbers <- matrix(NA, ncol=length( startyr:2017), nrow=NSim, dimnames=list("sim"=1:NSim,"year"=startyr:2017))

#we will need to extract correct group from spatiotemporal w. Those are numbered 1:xxx.
# we make a separate counter names gg starting at 1
gg <- 1

for (ii in startyr:2017){
for (jj in 1:NSim){
#intercept  +           year        +  ( 0 times beta for lsurface (so we get 1 km))
if (ii==startyr){
lin <- linmat["Intercept",jj] +                                 0 * linmat["lsurface",jj]
}else{
lin <- linmat["Intercept",jj] + linmat[paste0("fYear",ii),jj] + 0 * linmat["lsurface",jj]
}

# section below is for including spatial corr. Now we have to select the group. 
wm.pm1km2  <- inla.mesh.project(wproj, wmat[w.st$w.group==gg,jj])

# merging depth in?
dd <- approx(x=unique(cpue_subset$Depth),y=dmat[,jj], xout=c(depths))$y
dm.pm1km2 <-  matrix(dd, nrow=length(wproj$x),ncol=length(wproj$y), dimnames=list("x"=wproj$x,
"y"=wproj$y))

#only take dm and wm values only where wm.pm1km2 non-NAs
dm.pm1km2 <- dm.pm1km2[!is.na(wm.pm1km2)]
wm.pm1km2 <- wm.pm1km2[!is.na(wm.pm1km2)]

res <- exp(lin + wm.pm1km2 + dm.pm1km2)
numbers[jj,as.character(ii)] <- sum(rnbinom(n = length(res), mu = res, size = I4nb$summary.hyperpar$mean[1]))
}
gg <- gg+1
}
```

Number is the estimated number for the entire surface of the grid
what are the conf bounds for numbers per year?

```{r, eval=T, echo=T}
qnumbers <- apply(numbers,2, quantile,c(0.025,0.5,0.975))

plot(qnumbers[2,]/1e6, xlab="Years", ylab="Population numbers (millions)", las=1, type="l", ylim=c(0,max(qnumbers)), xaxt='n', yaxs="i")
lines(qnumbers[1,]/1e6, lty=2)
lines(qnumbers[3,]/1e6, lty=2)
axis(1,at=1:length(startyr:2017), labels= startyr:2017)
```

What are the average weights per year?
To go from number to weight we need alpha and beta from growth curve (for length in cm and resulting W is in g )
alpha=0.156650
beta=2.190 ref is Bedford et al. 1986

```{r, eval=T, echo=T}
alpha <- 0.156650
beta <- 2.190
full_cpue$wt <-  (alpha * (full_cpue$LngtClass/10)^beta * full_cpue$NoPerHaul) /1000
annualmnwt <- aggregate(cbind(wt,NoPerHaul) ~ Year, data= full_cpue,  FUN= "sum") 
annualmnwt <- within(annualmnwt, mnwt <- wt/NoPerHaul)

plot(mnwt~Year, data=annualmnwt, type="b", ylab="Mean weight (kg)", ylim=c(0,max(mnwt, na.rm = T)),las=1)
```

Multiply weights by population numbers to get estimated population biomass. Divide by 1e6 to go from kg to 1000 tonnes. 

```{r, eval=T, echo=T}
qwts <- annualmnwt[annualmnwt$Year >= startyr,]$mnwt * qnumbers / 1e6
```


```{r, eval=T, echo=T}
plot(qwts[2,], xlab="Years", ylab="Stock Biomass (1000 tonnes)", las=1, type="l", ylim=c(0,max(qwts)), xaxt='n',yaxs="i")
lines(qwts[1,], lty=2)
lines(qwts[3,], lty=2)
axis(1,at=1:length(startyr:2017), labels= startyr:2017)
```


## Simlulate data by sampling from posterior, but now from simpler model (I2nb; without depth and only spatial corr)


```{r, eval=T, echo=T}

# Simulate from model I1nb
set.seed(1234)
NSim <- 1000
Sim <- inla.posterior.sample(n = NSim, result = I2nb)
```

In this example, we thus sample `r NSim` times. The Sim object is a list of length NSim. Each element of  this list contains a single realization. Let's have a closer look.

```{r, eval=T, echo=T}
names(Sim[[1]])

#get out names of different types of rows in latent
rnames <- rownames(Sim[[1]]$latent)
rtypes <- unique(unlist(lapply(strsplit(rnames,":"),function(x){x[1]})))
rtypes
```

```{r, eval=T, echo=T}
#to get the w par realizations out: 
rtypes[3]
wrownum <- grep(paste0("^",rtypes[3]),rownames(Sim[[1]]$latent))
wmat <- sapply(Sim, function(x) {x$latent[wrownum]})
dim(wmat)

#to get the fixed effect par realizations out 
fixed <- rtypes[-(1:3)]
lrownum <- unlist(lapply(fixed, function(x) {grep(x, rownames(Sim[[1]]$latent), fixed = TRUE)}    ))
linmat <- sapply(Sim, function(x) {x$latent[lrownum]})
dim(linmat)
dimnames(linmat)[[1]] <- fixed
```


```{r, eval=T, echo=T}
numbers <- matrix(NA, ncol=length( startyr:2017), nrow=NSim, dimnames=list("sim"=1:NSim,"year"=startyr:2017))

#we will need to extract correct group from spatiotemporal w. Those are numbered 1:xxx.
# we make a separate counter names gg starting at 1

for (ii in startyr:2017){
  for (jj in 1:NSim){
    #intercept  +           year        +  ( 0 times beta for lsurface (so we get 1 km))
    if (ii==startyr){
      lin <- linmat["Intercept",jj] +                                 0 * linmat["lsurface",jj]
    }else{
      lin <- linmat["Intercept",jj] + linmat[paste0("fYear",ii),jj] + 0 * linmat["lsurface",jj]
    }
    
    # section below is for including spatial corr. Now we have to select the group. 
    wm.pm1km2  <- inla.mesh.project(wproj, wmat[,jj])
    
    #only take wm values only where wm.pm1km2 non-NAs
    wm.pm1km2 <- wm.pm1km2[!is.na(wm.pm1km2)]
    
    res <- exp(lin + wm.pm1km2 )
    numbers[jj,as.character(ii)] <- sum(rnbinom(n = length(res), mu = res, size = I2nb$summary.hyperpar$mean[1]))
  }
}
```

Number is the estimated number for the entire surface of the grid
what are the conf bounds for numbers per year?
  
  ```{r, eval=T, echo=T}
qnumbers <- apply(numbers,2, quantile,c(0.025,0.5,0.975))

plot(qnumbers[2,]/1e6, xlab="Years", ylab="Population numbers (millions)", las=1, type="l", ylim=c(0,max(qnumbers)), xaxt='n', yaxs="i")
lines(qnumbers[1,]/1e6, lty=2)
lines(qnumbers[3,]/1e6, lty=2)
axis(1,at=1:length(startyr:2017), labels= startyr:2017)
```

```{r, eval=T, echo=T}
qwts <- annualmnwt[annualmnwt$Year >= startyr,]$mnwt * qnumbers / 1e6
```

```{r, eval=T, echo=T}
plot(qwts[2,], xlab="Years", ylab="Stock Biomass (1000 tonnes)", las=1, type="l", ylim=c(0,max(qwts)), xaxt='n')
lines(qwts[1,], lty=2)
lines(qwts[3,], lty=2)
axis(1,at=1:length(startyr:2017), labels= startyr:2017)
```

Save image
```{r, eval=T, echo=T}
save.image("INLA_results2.Rdata")
```
